---
title: "analaysis"
output: html_document
---

```{r library, comment=NA, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(here)
```

### Data Wrangling  

These are all the datasets we obtained from the collaborators, which are

- BIOCARD_Demographics_2024.08.07_Deidentified.xlsx

- BIOCARD_DiagnosisData_2024.09.08_Deidentified.xlsx

- BIOCARD_Genetics_Data_2023.03.28_Deidentified.xlsx

- BIOCARD_Leisure_Activity_Champs_2023.09.10_other coded_Deidentified.xlsx

- BIOCARD_NTK Blood biomarkers_08.09.24_JHU only.xlsx

- BIOCARD_Vital_Signs_Sensory_2023.04.17_Deidentified.xlsx

- New participants_BIOCARD ApoE Genotypes 2023-2024_Deidentified.xlsx

- actigraphy_data_BIOCARD_8_7_24_from Daniel_Deidentified.xlsx

```{r data load, comment=NA, message=FALSE, warning=FALSE}
biocard_demographics <- read_xlsx("BIOCARD_Demographics_2024.08.07_Deidentified.xlsx")
biocard_diagnosis <- read_xlsx("BIOCARD_DiagnosisData_2024.09.08_Deidentified.xlsx")
biocard_genetics <- read_xlsx("BIOCARD_Genetics_Data_2023.03.28_Deidentified.xlsx")
biocard_apoe <- read_xlsx("New participants_BIOCARD ApoE Genotypes 2023-2024_Deidentified.xlsx")
biocard_leisure <- read_xlsx("BIOCARD_Leisure_Activity_Champs_2023.09.10_other coded_Deidentified.xlsx")
biocard_ntk_blood_biomarker <- read_xlsx("BIOCARD_NTK Blood biomarkers_08.09.24_JHU only.xlsx")
biocard_vital_signs_sensory <- read_xlsx("BIOCARD_Vital_Signs_Sensory_2023.04.17_Deidentified.xlsx")
biocard_actigraphy <- read_xlsx("actigraphy_data_BIOCARD_8_7_24_from Daniel_Deidentified.xlsx")
```

### Data Cleaning

We cleaned each of the data files based on the relevance to our analyses.

#### Demographics

```{r demographic, comment=NA, message=FALSE}
biocard_demographics_new <- biocard_demographics |>
    mutate(Sex_F = ifelse(SEX == 2, 1, 0),
           Race_White = ifelse(RACE == 1, 1, 0)) |>
    select(SUBJECT_ID, Sex_F, EDUC, Race_White)
head(biocard_demographics_new)
```

#### Diagnosis

```{r diagnosis, comment=NA, message=FALSE, warning=FALSE}
biocard_diagnosis_new <- biocard_diagnosis |>
    mutate(VISITNO = as.numeric(VISITNO)) |>
    arrange(SUBJECT_ID, VISITNO) |>
    group_by(SUBJECT_ID) |>
    mutate(fup_Dx = ifelse(DIAGNOSIS[n()] %in% c("MCI", "DEMENTIA"), 1, 0)) |> 
    ungroup() |>
    select(SUBJECT_ID, VISITNO, AgeAtVisit, DIAGNOSIS, DECAGE, fup_Dx) |>
    rename(AgeAtVisit_CHAMPS = AgeAtVisit)
head(biocard_diagnosis_new)
```

#### NTK Blood Biomarkers

```{r biomarker, comment=NA, message=FALSE, warning=FALSE}
biocard_ntk_blood_biomarker_new <- biocard_ntk_blood_biomarker |>
    select(SUBJECT_ID, VISITNO,
           PTAU181_zscore, AB42AB40_zscore, # AD-specific measures
           NFL_zscore, YKL40_zscore, sTREM2_zscore, GFAP_zscore, # AD non-specific measures
           PTAU181_outlier, AB42AB40_outlier, NFL_outlier,# outliers indicator
           YKL40_outlier, sTREM2_outlier, GFAP_outlier,
           PTAU181, PTAU181_AB42AB40, NFL, AB42AB40, YKL40, sTREM2, GFAP) |>
    mutate(VISITNO = as.numeric(VISITNO))
head(biocard_ntk_blood_biomarker_new)
```

#### CHAMPS activity questionnaire

```{r champs questionnaire, comment=NA, message=FALSE, warning=FALSE}
biocard_leisure_new <- biocard_leisure |>
    mutate(VISITNO = as.numeric(VISITNO),
           B11141s_CAT_low = ifelse(B11141s_CAT == "Physical-low", B11141a_freq, 0),
           B11141s_CAT_high = ifelse(B11141s_CAT == "Physical-modhi", B11141a_freq, 0)) |>
    group_by(SUBJECT_ID, VISITNO) |>
    mutate(LOW_INT_FREQ = sum(B11110a_freq, B11120a_freq, B11122a_freq, B11127a_freq, 
              B11128a_freq, B11134a_freq, B11135a_freq, B11139a_freq, B11141s_CAT_low, na.rm = TRUE),
           HIGH_INT_FREQ = sum(B11107a_freq, B11109a_freq, B11114a_freq, B11115a_freq,
              B11116a_freq, B11119a_freq, B11121a_freq, B11123a_freq, B11124a_freq,
              B11125a_freq, B11126a_freq, B11129a_freq, B11130a_freq, B11131a_freq,
              B11132a_freq, B11133a_freq, B11136a_freq, B11137a_freq, B11138a_freq, 
              B11140a_freq, B11141s_CAT_high, na.rm = TRUE),
           ALL_INT_FREQ = LOW_INT_FREQ + HIGH_INT_FREQ) |>
    select(SUBJECT_ID, VISITNO, LOW_INT_FREQ, HIGH_INT_FREQ, ALL_INT_FREQ)
head(biocard_leisure_new)
```

#### Actigraphy

```{r actigraphy, comment=NA, message=FALSE, warning=FALSE}
biocard_actigraphy_new <- biocard_actigraphy |>
    mutate(VISITNO = as.numeric(VISITNO)) |>
    group_by(SUBJECT_ID) |>
    # generate baseline variable for actigraphy assessment
    mutate(Age_Act = AgeAtVisit[act_visit == 1][1]) |> 
    rename("AgeAtVisit_Actigraphy" = "AgeAtVisit") |>
    select(SUBJECT_ID, VISITNO, Age_Act, AgeAtVisit_Actigraphy, LTAC10)
head(biocard_actigraphy_new)
```

#### Genetics

```{r genetics, comment=NA, message=FALSE, warning=FALSE}
biocard_apoe_1 <- biocard_apoe |>
    mutate(APOECODE = round(APOECODE/10,2))
biocard_genetics_new <- rbind(biocard_genetics, biocard_apoe_1) |>
    mutate(APOE4 = APOECODE,
           APOE4 = ifelse(APOECODE %in% c(2.2, 2.3, 3.3), 0, APOE4),
           APOE4 = ifelse(APOECODE %in% c(2.4, 3.4, 4.4), 1, APOE4)) |>
    select(-APOECODE)
head(biocard_genetics_new)    
```


#### Vital Signs/Sensory 

```{r sensory, comment=NA, message=FALSE, warning=FALSE}
biocard_vital_signs_sensory_new <- biocard_vital_signs_sensory |>
    select(SUBJECT_ID, VISITNO, BMI) |>
    mutate(VISITNO = as.numeric(VISITNO))
head(biocard_vital_signs_sensory_new)
```

### Merge datasets

```{r data merge, comment=NA, message=FALSE, warning=FALSE}
datafile_list <- list(biocard_diagnosis_new, biocard_demographics_new,
                      biocard_ntk_blood_biomarker_new, biocard_leisure_new,
                      biocard_genetics_new, biocard_actigraphy_new,
                      biocard_vital_signs_sensory_new)

merge_dfs <- function(df1, df2) {
  common_cols <- intersect(names(df1), names(df2))
  if ("VISITNO" %in% common_cols) {
    full_join(df1, df2, by = c("SUBJECT_ID", "VISITNO"))
  } else {
    full_join(df1, df2, by = "SUBJECT_ID")
  }
}

final_data <- reduce(datafile_list, merge_dfs) # 4527, 29
dim(final_data)
```

### Exclusion Criteria

```{r exclusion criteria, comment=NA, message=FALSE, warning=FALSE}
final_data_1 <- final_data |>
    # keep only JHU assessments
    filter(VISITNO > 100) |>
    # generate baseline variable for CHAMPS assessment
    group_by(SUBJECT_ID) |>
    mutate(Age_CHAMPS = AgeAtVisit_CHAMPS[VISITNO == 101][1]) |>
    ungroup() |>
    # reorder the columns
    select(SUBJECT_ID, VISITNO, DIAGNOSIS, AgeAtVisit_CHAMPS, Age_CHAMPS,
           AgeAtVisit_Actigraphy, Age_Act, everything())
dim(final_data_1) # 3099, 30

final_data_2 <- final_data_1 |>
    # participants with MCI or Dementia at their first available CHAMPS measure
    group_by(SUBJECT_ID) |>
    mutate(participant_move = any(DIAGNOSIS %in% c("MCI", "DEMENTIA") & AgeAtVisit_CHAMPS == Age_CHAMPS)) |> 
    ungroup() |> 
    filter(!participant_move) |>
    select(-participant_move)
dim(final_data_2) # 2593, 30

# at least one assessment within 1.5 years of baseline
final_data_3 <- final_data_2 |>
    mutate(age_diff = AgeAtVisit_CHAMPS - Age_CHAMPS) |>
    group_by(SUBJECT_ID) |>
    mutate(assessments_within_1_5_years = sum(age_diff >= 0 & age_diff <= 1.5, na.rm = TRUE)) |>
    ungroup() |>
    filter(assessments_within_1_5_years >= 1) |>
    select(-c(assessments_within_1_5_years, age_diff))
dim(final_data_3) # 2593, 30

# find subject id without zscores for 101 but zscores for 102
final_data_4_no_biomarker <- final_data_3 |>
    mutate(age_1_5 = Age_CHAMPS + 1.5) |> 
    filter(AgeAtVisit_CHAMPS <= age_1_5) |>
    filter(VISITNO %in% c(101, 102)) |>
    select(SUBJECT_ID, VISITNO, ends_with("zscore")) |>
    group_by(SUBJECT_ID) |>
    mutate(all_na_101 = all(is.na(across(ends_with("zscore")))[VISITNO == 101]),
           all_exist_102 = all(!is.na(across(ends_with("zscore")))[VISITNO == 102])) |>
    filter(all_na_101 & all_exist_102) |>
    mutate(n_subject = n()) |>
    ungroup() |>
    filter(n_subject > 1) 

subject_no_biomarker_101 <- unique(final_data_4_no_biomarker$SUBJECT_ID)
subject_no_biomarker_101 # 198 143 108 293 318 232  19

# copy 102 zscores to 101 scores
for (subject in subject_no_biomarker_101) {
    idx_101 <- which(final_data_3$SUBJECT_ID == subject & final_data_3$VISITNO == 101)
    idx_102 <- which(final_data_3$SUBJECT_ID == subject & final_data_3$VISITNO == 102)

    if (length(idx_101) > 0 & length(idx_102) > 0) {
        final_data_3[idx_101, grepl("zscore$", colnames(final_data_3))] <- 
            final_data_3[idx_102, grepl("zscore$", colnames(final_data_3))]
    }
    
    if (length(idx_101) > 0 & length(idx_102) > 0) {
        final_data_3[idx_101, grepl("outlier$", colnames(final_data_3))] <- 
            final_data_3[idx_102, grepl("outlier$", colnames(final_data_3))]
    }
} 
    
# remove outlier biomarkers
biomarkers <- c("PTAU181","AB42AB40","NFL","YKL40","sTREM2","GFAP")
final_data_4 <- final_data_3 |>
    filter(across(all_of(paste0(biomarkers, "_outlier")), ~ . != 1, .names = "outlier_{col}"))
dim(final_data_4) # 1363, 30

# remove visits without any biomarker information
final_data_5 <- final_data_4 |>
    filter(!(is.na(PTAU181_zscore) | is.na(AB42AB40_zscore) | is.na(NFL_zscore) |
       is.na(YKL40_zscore) | is.na(sTREM2_zscore) | is.na(GFAP_zscore)))
dim(final_data_5) # 1363, 30

length(unique(final_data_5$SUBJECT_ID))
```

## CHAMPS Analysis

```{r data champs, comment=NA, message=FALSE, warning=FALSE}
champs_data <- final_data_5
```

### Descriptive Analysis

```{r descriptive champs, comment=NA, message=FALSE, warning=FALSE}
length(unique(champs_data$SUBJECT_ID))
# 209

champs_data |>
    filter(DIAGNOSIS %in% c("MCI", "DEMENTIA") & fup_Dx == 1) |>
    nrow()
# 85

champs_data |>
    filter(SUBJECT_ID >= 400) |>
    nrow()
# 0

mean(champs_data$Age_CHAMPS) # 64.83
sd(champs_data$Age_CHAMPS) # 8.32
min(champs_data$Age_CHAMPS) # 28.22
max(champs_data$Age_CHAMPS) # 92.82

champs_data |>
    group_by(SUBJECT_ID) |>
    filter(Sex_F == 1) |>
    distinct(SUBJECT_ID) |>
    nrow()
# 128 (128/209 = 61.2%)

mean(champs_data$EDUC) # 17.44
sd(champs_data$EDUC) # 2.198

champs_data |>
    group_by(SUBJECT_ID) |>
    filter(Race_White == 1) |>
    distinct(SUBJECT_ID) |>
    nrow()
# 208 (208/209 = 99.5%)

champs_data |>
    group_by(SUBJECT_ID) |>
    filter(APOE4 == 1) |>
    distinct(SUBJECT_ID) |>
    nrow()
# 65 (65/209 = 31.1%)

follow_up_data <- champs_data |> 
    group_by(SUBJECT_ID) |>   
    filter(!is.na(Age_CHAMPS)) |>
    filter(row_number() == which.max(VISITNO)) |> 
    mutate(follow_up_time = AgeAtVisit_CHAMPS - Age_CHAMPS) |>
    select(SUBJECT_ID, VISITNO, Age_CHAMPS, AgeAtVisit_CHAMPS, follow_up_time) |> 
    ungroup()
mean(follow_up_data$follow_up_time) # 7.85
sd(follow_up_data$follow_up_time) # 2.77
min(follow_up_data$follow_up_time) # 0
max(follow_up_data$follow_up_time) # 11.72

mean(champs_data$LOW_INT_FREQ, na.rm = T) # 12.29
sd(champs_data$LOW_INT_FREQ, na.rm = T) # 8.29

mean(champs_data$HIGH_INT_FREQ, na.rm = T) # 9.98
sd(champs_data$HIGH_INT_FREQ, na.rm = T) # 8.82

mean(champs_data$ALL_INT_FREQ, na.rm = T) # 22.27
sd(champs_data$ALL_INT_FREQ, na.rm = T) # 13.11

mean(champs_data$BMI, na.rm = T) # 26.28
sd(champs_data$BMI, na.rm = T) # 4.72

baseline <- champs_data |> 
    filter(AgeAtVisit_CHAMPS == Age_CHAMPS)

mean(baseline$GFAP, na.rm = T) # 74.597
sd(baseline$GFAP, na.rm = T) # 34.43

mean(baseline$NFL, na.rm = T) # 2.49
sd(baseline$NFL, na.rm = T) # 1.25

mean(baseline$PTAU181, na.rm = T) # 0.711
sd(baseline$PTAU181, na.rm = T) # 0.227

mean(baseline$AB42AB40, na.rm = T) # 0.125
sd(baseline$AB42AB40, na.rm = T) # 0.0176

mean(baseline$PTAU181_AB42AB40, na.rm = T) # 5.865
sd(baseline$PTAU181_AB42AB40, na.rm = T) # 2.32

mean(baseline$sTREM2, na.rm = T) # 6711.27
sd(baseline$sTREM2, na.rm = T) # 2829.39

mean(baseline$YKL40, na.rm = T) # 65891.53
sd(baseline$YKL40, na.rm = T) # 71101.58
```


## Actigraphy Analysis

```{r data actigraphy, comment=NA, message=FALSE, warning=FALSE }
actigraphy_data <- final_data_5 |>
  filter(!is.na(AgeAtVisit_Actigraphy))
```

### Descriptive

```{r descriptive actigraphy, comment=NA, message=FALSE, warning=FALSE}
length(unique(actigraphy_data$SUBJECT_ID))
# 150

actigraphy_data |>
    filter(DIAGNOSIS %in% c("MCI", "DEMENTIA") & fup_Dx == 1) |>
    nrow()
# 22

actigraphy_data |>
    filter(SUBJECT_ID >= 400) |>
    nrow()
# 0

mean(actigraphy_data$Age_Act, na.rm = T) # 70.989
sd(actigraphy_data$Age_Act, na.rm = T) # 8.026
min(actigraphy_data$Age_Act, na.rm = T) # 39.329
max(actigraphy_data$Age_Act, na.rm = T) # 93.25

actigraphy_data |>
    group_by(SUBJECT_ID) |>
    filter(Sex_F == 1) |>
    distinct(SUBJECT_ID) |>
    nrow()
# 95 (95/150 = 63.3%)

mean(actigraphy_data$EDUC) # 17.37
sd(actigraphy_data$EDUC) # 2.161

actigraphy_data |>
    group_by(SUBJECT_ID) |>
    filter(Race_White == 1) |>
    distinct(SUBJECT_ID) |>
    nrow()
# 149 (149/150 = 99.3%)

actigraphy_data |>
    group_by(SUBJECT_ID) |>
    filter(APOE4 == 1) |>
    distinct(SUBJECT_ID) |>
    nrow()
# 47 (47/150 = 31.3%)

follow_up_data_actigraphy <- actigraphy_data |> 
    group_by(SUBJECT_ID) |>   
    filter(!is.na(Age_Act)) |>
    filter(row_number() == which.max(VISITNO)) |> 
    mutate(follow_up_time = AgeAtVisit_Actigraphy - Age_Act) |>
    select(SUBJECT_ID, VISITNO, Age_Act, AgeAtVisit_Actigraphy, follow_up_time) |> 
    ungroup()
mean(follow_up_data_actigraphy$follow_up_time, na.rm = T) # 1.1487
sd(follow_up_data_actigraphy$follow_up_time, na.rm = T) # 1.098
min(follow_up_data_actigraphy$follow_up_time, na.rm = T) # 0
max(follow_up_data_actigraphy$follow_up_time, na.rm = T) # 3.428

mean(actigraphy_data$LOW_INT_FREQ, na.rm = T) # 12.818
sd(actigraphy_data$LOW_INT_FREQ, na.rm = T) # 7.844

mean(actigraphy_data$HIGH_INT_FREQ, na.rm = T) # 10.138
sd(actigraphy_data$HIGH_INT_FREQ, na.rm = T) # 8.335

mean(actigraphy_data$ALL_INT_FREQ, na.rm = T) # 22.96
sd(actigraphy_data$ALL_INT_FREQ, na.rm = T) # 12.46

mean(actigraphy_data$BMI, na.rm = T) # 26.22
sd(actigraphy_data$BMI, na.rm = T) # 3.63

baseline <- actigraphy_data |> 
    filter(AgeAtVisit_Actigraphy == Age_Act)

mean(baseline$GFAP, na.rm = T) # 88.2957
sd(baseline$GFAP, na.rm = T) # 40.73335

mean(baseline$NFL, na.rm = T) # 2.980845
sd(baseline$NFL, na.rm = T) # 1.349039

mean(baseline$PTAU181, na.rm = T) # 0.8435352
sd(baseline$PTAU181, na.rm = T) # 0.3064263

mean(baseline$AB42AB40, na.rm = T) # 0.1277212
sd(baseline$AB42AB40, na.rm = T) # 0.01819899

mean(baseline$PTAU181_AB42AB40, na.rm = T) # 6.84814
sd(baseline$PTAU181_AB42AB40, na.rm = T) # 2.999189

mean(baseline$sTREM2, na.rm = T) # 7716.972
sd(baseline$sTREM2, na.rm = T) # 3151.575

mean(baseline$YKL40, na.rm = T) # 76638.52
sd(baseline$YKL40, na.rm = T) # 70955.52
```
